---
title: "Hands-on Exercise 8"
author: "Cheok En Qi"
date: "March 1, 2026"
date-modified: "last-modified"
format: html
editor: visual
---

# Lesson 8a: Choropleth Mapping with R

## Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Data

### 21.3.2 Importing Geospatial Data into R

The code chunk below uses the st_read() function of sf package to import MP14_SUBZONE_WEB_PL shapefile into R as a simple feature data frame called mpsz.

```{r}
mpsz <- st_read(dsn = "chap08/data", layer = "MP14_SUBZONE_WEB_PL")

mpsz
```

### 21.3.3 Importing Attribute Data into R

```{r}
popdata <- read_csv("chap08/data/respopagesextod2011to2020.csv")
```

## 21.3.4 Data Preparation

Prepare a data table with year 2020 values. The data table YOUNG: age group 0 to 4 until age groyup 20 to 24, ECONOMY ACTIVE: age group 25-29 until age group 60-64, AGED: age group 65 and above, TOTAL: all age group, and DEPENDENCY: the ratio between young and aged against economy active group

### 21.3.4.1 Data wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### 21.3.4.2 Joining the attribute data and geospatial data

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)

mpsz_pop2020 <- left_join(mpsz, popdata2020, by = c("SUBZONE_N" = "SZ"))

write_rds(mpsz_pop2020, "chap08/data/mpszpop2020.rds")
```

## 21.4 Choropleth Mapping Geospatial Data Using tmap

### 21.4.1 Plotting a choropleth map quickly by using qtm()

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

### 21.4.2 Creating a choropleth map by using tmap’s elements

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.blues"),fill.legend = tm_legend(title = "Dependency ratio")) + tm_title("Distribution of Dependency Ratio by planning subzone") + tm_layout(frame = TRUE) + tm_borders(fill_alpha = 0.5) + tm_compass(type="8star", size = 2) + tm_grid(alpha =0.2) + tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", position = c("left", "bottom"))
```

#### 21.4.2.1 Drawing a base map

```{r}
tm_shape(mpsz_pop2020) + tm_polygons()
```

#### 21.4.2.2 Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY")
```

####21.4.2.3 Drawing a choropleth map using tm_fill() and \*tm_border()\*\*

```{r}
tm_shape(mpsz_pop2020)+ tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY") + tm_borders(lwd = 0.01, fill_alpha = 0.1)
```

### 21.4.3 Data classification methods of tmap

#### 21.4.3.1 Plotting choropleth maps with built-in classification methods

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "jenks", n = 5)) + tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "equal", n = 5)) + tm_borders(fill_alpha = 0.5)
```

#### 21.4.3.2 Plotting choropleth map with custome break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) + tm_borders(fill_alpha = 0.5)
```

### 21.4.4 Colour Scheme

#### 21.4.4.1 Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.greens")) + tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "-brewer.greens")) + tm_borders(fill_alpha = 0.5)
```

### 21.4.5 Map Layouts

#### 21.4.5.1 Map Legend

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "jenks", n = 5, values = "brewer.greens"), fill.legend = tm_legend(title = "Dependency ratio")) + tm_borders(fill_alpha = 0.5) + tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

#### 21.4.5.2 Map style

```{r}
tm_shape(mpsz_pop2020)+ tm_fill("DEPENDENCY", style = "quantile", palette = "-Greens") + tm_borders(alpha = 0.5) + tmap_style("classic")
```

#### 21.4.5.3 Cartographic Furniture

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.blues"), fill.legend = tm_legend(title = "Dependency ratio")) + tm_title("Distribution of Dependency Ratio by planning subzone") + tm_layout(frame = TRUE) + tm_borders(fill_alpha = 0.5) + tm_compass(type="8star", size = 2) + tm_grid(alpha =0.2) + tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", position = c("left", "bottom"))


tmap_style("white")
```

### 21.4.6 Drawing Small Multiple Choropleth Maps

#### 21.4.6.1 By assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ tm_fill(c("YOUNG", "AGED"), style = "equal", palette = "Blues") + tm_layout(legend.position = c("right", "bottom")) + tm_borders(alpha = 0.5) + tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(c("DEPENDENCY","AGED"), style = c("equal", "quantile"), palette = list("Blues","Greens")) + tm_layout(legend.position = c("right", "bottom"))
```

#### 21.4.6.2 By defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) + tm_fill("DEPENDENCY", style = "quantile",
 palette = "Blues", thres.poly = 0) + tm_facets(by="REGION_N",  free.coords=TRUE) + tm_layout(legend.show = FALSE, title.position = c("center", "center"), title.size = 20) + tm_borders(alpha = 0.5)
```

#### 21.4.6.3 By creating multiple stand-alone maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ tm_polygons("YOUNG", style = "quantile", palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ tm_polygons("AGED", style = "quantile", palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

#### 21.4.7 Mappping Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
tm_fill("DEPENDENCY", style = "quantile", palette = "Blues", legend.hist = TRUE, legend.is.portrait = TRUE, legend.hist.z = 0.1) + tm_layout(legend.outside = TRUE, legend.height = 0.45, legend.width = 5.0, legend.position = c("right", "bottom"), frame = FALSE) + tm_borders(alpha = 0.5)
```

# Lesson 8b: Visualising Geospatial Point Data

## Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Data

Import dataset..

```{r}
sgpools <- read_csv("chap08/data/SGPools_svy21.csv")
```

```{r}
list(sgpools) 
```

## 25.3 Creating a sf data frame from an aspatial data frame

```{r}
sgpools_sf <- st_as_sf(sgpools, coords = c("XCOORD", "YCOORD"), crs= 3414)
```

```{r}
list(sgpools_sf)
```

## 26 Drawing Proportional Symbol Map

```{r}
tmap_mode("view")
```

### 26.1 It all started with an interactive point symbol map

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "red", size = 1, col = "black", lwd = 1)
```

### 26.2 Lets make it proportional

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "red", size = "Gp1Gp2 Winnings", col = "black", lwd = 1)
```

### 26.3 Lets give it a different colour

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "OUTLET TYPE", size = "Gp1Gp2 Winnings", col = "black", lwd = 1)
```

### 26.4 I have a twin brothers - create twin map

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "OUTLET TYPE", size = "Gp1Gp2 Winnings", col = "black", lwd = 1) + tm_facets(by= "OUTLET TYPE", nrow = 1, sync = TRUE)
```

```{r}
tmap_mode("plot")
```

# Lesson 8c: Analytical Mapping

## Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Data

Import dataset

```{r}
NGA_wp <- read_rds("chap08/data/NGA_wp.rds")
```

## 23.3 Basic Choropleth Mapping

### 23.3.1 Visualising distribution of non-functional water point

```{r}
p1 <- tm_shape(NGA_wp) + tm_polygons(fill = "wp_functional", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) + tm_title("Distribution of functional water point by LGAs")
```

```{r}
p2 <- tm_shape(NGA_wp) + tm_polygons(fill = "total_wp", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) +  tm_title("Distribution of total  water point by LGAs")
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

## 23.4 Choropleth Map for Rates

### 23.4.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### 23.4.2 Plotting map of rate

```{r}
tm_shape(NGA_wp) + tm_polygons("pct_functional", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) + tm_title("Rate map of functional water point by LGAs")
```

## 23.5 Extreme Value Maps

### 23.5.1 Percentile Map

#### 23.5.1.1 Data Preparation

Step 1: Exclude records with NA by using the code chunk below

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

#### 23.5.1.2 Why writing functions?

#### 23.5.1.3 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### 23.5.1.4 A percentile mapping function

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){percent <- c(0,.01,.1,.5,.9,.99,1)
var <- get.var(vnam, df)
bperc <- quantile(var, percent)
tm_shape(df) + tm_polygons() + tm_shape(df) + tm_polygons(vnam, title=legtitle, breaks=bperc, palette="Blues", labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  + tm_borders() + tm_layout(main.title = mtitle, title.position = c("right","bottom"))
}
```

#### 23.5.1.5 Test drive the percentile mapping function

```{r}
percentmap("total_wp", NGA_wp)
```

### 23.5.2 Box map

```{r}
ggplot(data = NGA_wp, aes(x = "", y = wp_nonfunctional)) + geom_boxplot()
```

#### 23.5.2.1 Creating the boxbreaks function

```{r}
boxbreaks <- function(v,mult=1.5) {
qv <- unname(quantile(v))
iqr <- qv[4] - qv[2]
upfence <- qv[4] + mult * iqr
lofence <- qv[2] - mult * iqr
# initialize break points vector
bb <- vector(mode="numeric",length=7)
# logic for lower and upper fences
if (lofence < qv[1]) {  # no lower outliers
  bb[1] <- lofence
  bb[2] <- floor(qv[1])
} else {
  bb[2] <- lofence
  bb[1] <- qv[1]
}
if (upfence > qv[5]) { # no upper outliers
  bb[7] <- upfence
  bb[6] <- ceiling(qv[5])
} else {
  bb[6] <- upfence
  bb[7] <- qv[5]
}
bb[3:5] <- qv[2:4]
return(bb)
}
```

#### 23.5.2.2 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### 23.5.2.3 Test drive the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

#### 23.5.2.4 Boxmap function

```{r}
boxmap <- function(vnam, df, legtitle=NA, mtitle="Box Map", mult=1.5){
var <- get.var(vnam,df)
bb <- boxbreaks(var)
tm_shape(df) + tm_polygons() + tm_shape(df) + tm_fill(vnam,title=legtitle, breaks=bb, palette="Blues", labels = c("lower outlier", "< 25%", "25% - 50%", "50% - 75%", "> 75%", "upper outlier")) + tm_borders() + tm_layout(main.title = mtitle,  title.position = c("left", "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
