---
title: "Hands-on Exercise 9"
author: "Cheok En Qi"
date: "March 8, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 9: Modelling, Visualising and Analysing Network Data with R

##Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph, visNetwork, lubridate, clock, tidyverse, graphlayouts, concaveman, ggforce)
```

## Data

### Importing Data

```{r}
GAStech_nodes <- read_csv("chap09/data/GAStech_email_node.csv")
GAStech_edges <- read_csv("chap09/data/GAStech_email_edge-v2.csv")
```

### 27.3.3 Importing network data from files

```{r}
glimpse(GAStech_edges)
```

### 27.3.5 Wrangling time

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate, label = TRUE, abbr = FALSE))
```

### 27.3.7 Wrangling attributes

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## 27.4 Creating network objects using tidygraph
### 27.4.3 Using tbl_graph() to build tidygraph data model.

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes, edges = GAStech_edges_aggregated, directed = TRUE)
```

### 27.4.4 Reviewing the output tidygraph’s graph object

```{r}
GAStech_graph
```

### 27.4.6 Changing the active object

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## 27.5 Plotting Static Network Graphs with ggraph package

### 27.5.1 Plotting a basic network graph

```{r}
ggraph(GAStech_graph) + geom_edge_link() + geom_node_point()
```

### 27.5.2 Changing the default network graph theme

```{r}
g <- ggraph(GAStech_graph) + geom_edge_link(aes()) + geom_node_point(aes())

g + theme_graph()
```

### 27.5.3 Changing the coloring of the plot

```{r}
g <- ggraph(GAStech_graph) + geom_edge_link(aes(colour = 'grey50')) + geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10', text_colour = 'white')
```

### 27.5.5 Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, layout = "fr") + geom_edge_link(aes()) +
geom_node_point(aes())

g + theme_graph()
```

### 27.5.6 Modifying network nodes

```{r}
g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes()) + geom_node_point(aes(colour = Department,  size = 3))

g + theme_graph()
```

### 27.5.7 Modifying edges

```{r}
g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) +  scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 3)

g + theme_graph()
```

## 27.6 Creating facet graphs

### 27.6.1 Working with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)

g + facet_edges(~Weekday)
```

### 27.6.2 Working with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2) + theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

### 27.6.3 A framed facet graph

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, layout = "nicely") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)
  
g + facet_edges(~Weekday) + th_foreground(foreground = "grey80", border = TRUE) + theme(legend.position = 'bottom')
```

### 27.6.4 Working with facet_nodes()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)
  
g + facet_nodes(~Department)+ th_foreground(foreground = "grey80", border = TRUE) + theme(legend.position = 'bottom')
```

## 27.7 Network Metrics Analysis

### 27.7.1 Computing centrality indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department, size=betweenness_centrality))
g + theme_graph()
```

### 27.7.2 Visualising network metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department, size = centrality_betweenness()))
g + theme_graph()
```

### 27.7.3 Visualising Community

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(
aes(colour = community))  

g + theme_graph()
```

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(group_optimal(weights = Weight)),  betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") + geom_mark_hull(aes(x, y, group = community, fill = community), alpha = 0.2, expand = unit(0.3, "cm"), radius = unit(0.3, "cm")) + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(fill = Department, size = betweenness_measure), color = "black", shape = 21)
  
g + theme_graph()
```

## 27.8 Building Interactive Network Graph with visNetwork

### 27.8.1 Data preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### 27.8.2 Plotting the first interactive network graph

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated)
```

### 27.8.3 Working with layout

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

### 27.8.4 Working with visual attributes - Nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### 27.8.5 Working with visual attributes - Edges

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", smooth = list(enabled = TRUE, type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123) 
```

### 27.8.6 Interactivity

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
