---
title: "Take-Home Exercise 1:Survey Analysis of Informal Labourers' Income"
author: "Cheok En Qi"
date: "February 8, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

This report analyses the income of informal rural labourers in the northern mountainous regions of Vietnam.

------------------------------------------------------------------------

# Getting Started
Install and load the required R packages for data manipulation and visualisation.
```{r}
pacman::p_load(tidyverse, readxl, knitr, ggthemes, patchwork)
```

# Loading Data
The dataset was imported from an Excel file containing survey responses from informal rural labourers in the northern mountainous regions of Vietnam.
```{r}
df_raw <- read_xlsx("chapthe/data/Upload for elsiver.xlsx")
glimpse(df_raw)

```


# Data Preparation
To prepare the dataset for analysis and visualisation, several cleaning and recoding steps were performed.

1. Key categorical variables (province, gender, race, education level, vocational training type, and credit access) were converted into factor variables to support clearer grouped comparisons in plots.

2. Monetary and numeric variables (total income, savings, loans, and working days) were converted into numeric format to ensure accurate computation and visualisation.

3. Invalid data were recoded as missing values (NA) to reduce distortion in summary statistics. 

4. Total income was defined using tein, which represents total earning

```{r}
df <- df_raw %>%
 
  rename_with(tolower) %>% 
  

  mutate(
    cpro = factor(cpro,
                  levels = 1:5,
                  labels = c("Tuyen Quang","Quang Ninh","Ha Giang","Yen Bai","Bac Giang")),

    cgen = factor(cgen,
                  levels = c(1,2),
                  labels = c("Male","Female")),

    crac = factor(crac,
                  levels = c(1,2),
                  labels = c("King","Minorities")),

    cjob = factor(cjob,
                  levels = c(1,2,3),
                  labels = c("Agriculture","Service/Industrial","Others")),

    cqui = factor(cqui,
                  levels = c(1,2,3,4,5),
                  labels = c("Top","Fourth","Middle","Second","Lowest")),

    fedu = factor(fedu,
                  levels = c(0,1,2,3),
                  labels = c("Primary","Lower secondary","Upper secondary","Others")),

    fvtp = factor(fvtp,
                  levels = c(1,2),
                  labels = c("Short course","Long course")),

    fcra = factor(fcra,
                  levels = c(1,2),
                  labels = c("Inaccessible","Accessible")),

    ftap = factor(ftap,
                  levels = c(1,2),
                  labels = c("Low technique","Adequate")),

    tein = as.numeric(tein),
    tain = as.numeric(tain),
    tsii = as.numeric(tsii),
    toin = as.numeric(toin),
    lcre = as.numeric(lcre),
    lsav = as.numeric(lsav),
    lwda = as.numeric(lwda),

    income = tein
  ) %>% 

  filter(!is.na(income))


glimpse(df)
```

------------------------------------------------------------------------

# Visualisation 1: Income Distribution

------------------------------------------------------------------------

:::: callout-note
## Visualisation 1: Income Distribution

::: panel-tabset
### Visual

```{r, echo=FALSE}
ggplot(df %>% drop_na(income), aes(x = income)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    color = "black",
    fill = "lightblue",
    alpha = 0.6
  ) +
  geom_density(linewidth = 1) +
  labs(
    title = "Distribution of Total Income",
    x = "Income (Million VND)",
    y = "Density"
  ) +
  theme_minimal()

# Summary Table 
knitr::kable(
  df %>%
    summarise(
      Count = sum(!is.na(income)),
      Mean = mean(income, na.rm = TRUE),
      Median = median(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      Min = min(income, na.rm = TRUE),
      Q1 = quantile(income, 0.25, na.rm = TRUE),
      Q3 = quantile(income, 0.75, na.rm = TRUE),
      Max = max(income, na.rm = TRUE)
    ),
  digits = 2,
  caption = "Summary Statistics of Income"
)

```

### Code

``` r
ggplot(df %>% drop_na(income), aes(x = income)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    color = "black",
    fill = "lightblue",
    alpha = 0.6
  ) +
  geom_density(linewidth = 1) +
  labs(
    title = "Distribution of Total Income",
    x = "Income (Million VND)",
    y = "Density"
  ) +
  theme_minimal()

# Summary Table 
knitr::kable(
  df %>%
    summarise(
      Count = sum(!is.na(income)),
      Mean = mean(income, na.rm = TRUE),
      Median = median(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      Min = min(income, na.rm = TRUE),
      Q1 = quantile(income, 0.25, na.rm = TRUE),
      Q3 = quantile(income, 0.75, na.rm = TRUE),
      Max = max(income, na.rm = TRUE)
    ),
  digits = 2,
  caption = "Summary Statistics of Income"
)


```
:::

**Observation:**\
The income distribution is heavily right-skewed, with the majority of labourers earning in the lower brackets. The long tail on the right shows that a small number earn much more than the rest, highlighting uneven earnings and suggest that income is unevenly distributed. 

::::

------------------------------------------------------------------------

# Visualisation 2: Income Across Socioeconomic Factors (Province, Education, Credit)

------------------------------------------------------------------------

:::: callout-note
## Visualisation 2: Income Across Socioeconomic Factors (Province, Education, Credit)

::: panel-tabset
### Visual

```{r, echo=FALSE}
y_max <- quantile(df$income, 0.98, na.rm = TRUE) * 1.10


# Plot A: Income by Province
df_cpro <- df %>% drop_na(income, cpro)

p_cpro <- ggplot(df_cpro, aes(x = cpro, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Province",
    x = "Province",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#Plot B: Income by Education
df_fedu <- df %>% drop_na(income, fedu)

p_fedu <- ggplot(df_fedu, aes(x = fedu, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Education Level",
    x = "Education Level",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#  Plot C: Income by Credit Access
df_fcra <- df %>% drop_na(income, fcra)

p_fcra <- ggplot(df_fcra, aes(x = fcra, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Credit Access",
    x = "Credit Access",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

# Display the combined plots
p_cpro + p_fedu + p_fcra
```

### Code

``` r
y_max <- quantile(df$income, 0.98, na.rm = TRUE) * 1.10


# Plot A: Income by Province
df_cpro <- df %>% drop_na(income, cpro)

p_cpro <- ggplot(df_cpro, aes(x = cpro, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Province",
    x = "Province",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#Plot B: Income by Education
df_fedu <- df %>% drop_na(income, fedu)

p_fedu <- ggplot(df_fedu, aes(x = fedu, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Education Level",
    x = "Education Level",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#  Plot C: Income by Credit Access
df_fcra <- df %>% drop_na(income, fcra)

p_fcra <- ggplot(df_fcra, aes(x = fcra, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Credit Access",
    x = "Credit Access",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

# Display the combined plots
p_cpro + p_fedu + p_fcra

```
:::

**Observation:**\
Income varies across provinces, education levels, and credit access categories. Education shows the clearest pattern, with higher education levels generally associated with higher median incomes. Credit access is linked to slightly higher earnings, while provincial differences are present but less pronounced. 
::::

------------------------------------------------------------------------

# Visualisation 3: Income Across Demographic Characteristics (Gender, Training)

------------------------------------------------------------------------

:::: callout-note
## Visualisation 3: Income Across Demographic Characteristics (Gender, Training)

::: panel-tabset
### Visual

```{r, echo=FALSE}

y_max <- quantile(df$income, 0.95, na.rm = TRUE)

# Plot A: Income by Gender
df_gender <- df %>% drop_na(income, cgen)

p_gender <- ggplot(df_gender, aes(x = cgen, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Gender",
    x = "Gender",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#  Plot B: Income by Vocational Training Type
df_training <- df %>% drop_na(income, fvtp)

p_training <- ggplot(df_training, aes(x = fvtp, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Vocational Training Type",
    x = "Training type",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

p_gender + p_training

# 5. Summary Table
summary_v3 <- bind_rows(
  df_gender %>%
    group_by(Group = "Gender", Category = cgen) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    ),
  df_training %>%
    group_by(Group = "Training type", Category = fvtp) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    )
)

knitr::kable(summary_v3, digits = 2)
```

### Code

``` r

y_max <- quantile(df$income, 0.95, na.rm = TRUE)

# Plot A: Income by Gender
df_gender <- df %>% drop_na(income, cgen)

p_gender <- ggplot(df_gender, aes(x = cgen, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Gender",
    x = "Gender",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

#  Plot B: Income by Vocational Training Type
df_training <- df %>% drop_na(income, fvtp)

p_training <- ggplot(df_training, aes(x = fvtp, y = income)) +
  geom_boxplot(fill = "steelblue") +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(
    title = "Income by Vocational Training Type",
    x = "Training type",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

p_gender + p_training

# 5. Summary Table
summary_v3 <- bind_rows(
  df_gender %>%
    group_by(Group = "Gender", Category = cgen) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    ),
  df_training %>%
    group_by(Group = "Training type", Category = fvtp) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    )
)

knitr::kable(summary_v3, digits = 2)
```
:::

**Observation:**\
Income differences across gender appear minimal, with considerable overlap between the two groups. In contrast, vocational training type shows a clearer gap: respondents who attended long courses report substantially higher incomes than those who attended short courses. However, variability within each group remains large, especially for the long-course group. This suggests that factors beyond gender and training type also contribute to income differences.
::::

------------------------------------------------------------------------

# Visualisation 4: Association Between Income and Savings

------------------------------------------------------------------------

:::: callout-note
## Visualisation 4:Association Between Income and Savings

::: panel-tabset
### Visual

```{r, echo=FALSE}
df_scatter <- df %>% drop_na(income, lsav)

x_limit <- quantile(df_scatter$income, 0.98, na.rm = TRUE)
y_limit <- quantile(df_scatter$lsav,  0.98, na.rm = TRUE)

ggplot(df_scatter, aes(x = income, y = lsav)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  coord_cartesian(xlim = c(0, x_limit), ylim = c(0, y_limit)) +
  labs(
    title = "Association Between Income and Savings",
    x = "Income (Million VND)",
    y = "Savings"
  ) +
  theme_minimal()

# Summary table
knitr::kable(
  df_scatter %>%
    summarise(
      N = n(),
      Correlation_Pearson = cor(income, lsav, method = "pearson", use = "complete.obs"),
      Correlation_Spearman = cor(income, lsav, method = "spearman", use = "complete.obs"),
      Income_Mean = mean(income, na.rm = TRUE),
      Savings_Mean = mean(lsav, na.rm = TRUE)
    ),
  digits = 3,
  caption = "Summary of Income and Savings Relationship"
)
```

### Code

``` r
df_scatter <- df %>% drop_na(income, lsav)

x_limit <- quantile(df_scatter$income, 0.98, na.rm = TRUE)
y_limit <- quantile(df_scatter$lsav,  0.98, na.rm = TRUE)

ggplot(df_scatter, aes(x = income, y = lsav)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  coord_cartesian(xlim = c(0, x_limit), ylim = c(0, y_limit)) +
  labs(
    title = "Association Between Income and Savings",
    x = "Income (Million VND)",
    y = "Savings"
  ) +
  theme_minimal()

# Summary table
knitr::kable(
  df_scatter %>%
    summarise(
      N = n(),
      Correlation_Pearson = cor(income, lsav, method = "pearson", use = "complete.obs"),
      Correlation_Spearman = cor(income, lsav, method = "spearman", use = "complete.obs"),
      Income_Mean = mean(income, na.rm = TRUE),
      Savings_Mean = mean(lsav, na.rm = TRUE)
    ),
  digits = 3,
  caption = "Summary of Income and Savings Relationship"
)
``` 
:::

**Observation:**\
Savings show a weak relationship with income, with points widely dispersed and clustered at lower savings levels. A small number of extreme values influence the fitted trend line, and most respondents report relatively low savings even as income increases.
::::

------------------------------------------------------------------------


# Visualisation 5: Income by Perception Score (PPO1)

------------------------------------------------------------------------

:::: callout-note
## Visualisation 5: Income by Perception Score (PPO1)

::: panel-tabset
### Visual

```{r, echo=FALSE}
df_ppo1 <- df %>%
  mutate(ppo1 = na_if(ppo1, 0)) %>%
  drop_na(ppo1, income) %>%
  mutate(ppo1 = factor(ppo1, levels = 1:5, ordered = TRUE))

upper_limit <- quantile(df_ppo1$income, 0.95, na.rm = TRUE)

ggplot(df_ppo1, aes(x = ppo1, y = income)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, upper_limit)) +
  labs(
    title = "Income by PPO1 Score",
    x = "PPO1 score (1–5)",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

knitr::kable(
  df_ppo1 %>%
    group_by(ppo1) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    ),
  digits = 2,
  caption = "Income Summary by PPO1 Score<br>Scale: 1 = strongly disagree to 5 = strongly agree (as defined in the survey)"
)
```

### Code

``` r
df_ppo1 <- df %>%
  mutate(ppo1 = na_if(ppo1, 0)) %>%
  drop_na(ppo1, income) %>%
  mutate(ppo1 = factor(ppo1, levels = 1:5, ordered = TRUE))

upper_limit <- quantile(df_ppo1$income, 0.95, na.rm = TRUE)

ggplot(df_ppo1, aes(x = ppo1, y = income)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, upper_limit)) +
  labs(
    title = "Income by PPO1 Score",
    x = "PPO1 score (1–5)",
    y = "Income (Million VND)"
  ) +
  theme_minimal()

knitr::kable(
  df_ppo1 %>%
    group_by(ppo1) %>%
    summarise(
      N = n(),
      Median_Income = median(income, na.rm = TRUE),
      Mean_Income = mean(income, na.rm = TRUE),
      SD = sd(income, na.rm = TRUE),
      .groups = "drop"
    ),
  digits = 2,
  caption = "Income Summary by PPO1 Score<br>Scale: 1 = strongly disagree to 5 = strongly agree (as defined in the survey)"
)
```
:::

**Observation:**\
PPO1 reflects respondents’ views on training-related policy/support. This visualisation explores whether income levels differ across PPO1 score categories. Income distributions overlap heavily across PPO1 score levels, and median income changes only minimally from scores 1 to 5. While the mean income is higher for some scores , the large differences  indicate that a small number of high-income outliers likely drive these differences. Overall, the plot suggests no clear or strong relationship between PPO1 score and income.
::::

------------------------------------------------------------------------

# Visualisation 6: Heatmap of Income Distribution by Education Level

------------------------------------------------------------------------

:::: callout-note
## Visualisation 6: Heatmap of Income Distribution by Education Level

::: panel-tabset
### Visual

```{r, echo=FALSE}
df_hm <- df %>%
  drop_na(income, fedu) %>%
  mutate(
    
    income_band = ntile(income, 4),
    income_band = factor(
      income_band,
      levels = 1:4,
      labels = c("Q1 (Lowest)", "Q2", "Q3", "Q4 (Highest)"),
      ordered = TRUE
    )
  )

#  Plot Heatmap
df_hm %>%
  count(fedu, income_band) %>%
  group_by(fedu) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = fedu, y = income_band, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.5) +
 
  scale_fill_gradient(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Heatmap of Income Band by Education Level",
    x = "Education Level",
    y = "Income Band",
    fill = "% within Edu"
  ) +
  theme_minimal(base_size = 12)

# Summary Table
knitr::kable(
  df_hm %>%
    count(fedu, name = "Respondents") %>%
    mutate(Percentage = scales::percent(Respondents / sum(Respondents), accuracy = 0.1)),
  col.names = c("Education Level", "Respondents", "Percentage")
)

cat("Total number of respondents (used in heatmap):", nrow(df_hm))
```

### Code

``` r
df_hm <- df %>%
  drop_na(income, fedu) %>%
  mutate(
    
    income_band = ntile(income, 4),
    income_band = factor(
      income_band,
      levels = 1:4,
      labels = c("Q1 (Lowest)", "Q2", "Q3", "Q4 (Highest)"),
      ordered = TRUE
    )
  )

#  Plot Heatmap
df_hm %>%
  count(fedu, income_band) %>%
  group_by(fedu) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = fedu, y = income_band, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.5) +
 
  scale_fill_gradient(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Heatmap of Income Band by Education Level",
    x = "Education Level",
    y = "Income Band",
    fill = "% within Edu"
  ) +
  theme_minimal(base_size = 12)

# Summary Table
knitr::kable(
  df_hm %>%
    count(fedu, name = "Respondents") %>%
    mutate(Percentage = scales::percent(Respondents / sum(Respondents), accuracy = 0.1)),
  col.names = c("Education Level", "Respondents", "Percentage")
)

cat("Total number of respondents (used in heatmap):", nrow(df_hm))
```
:::

**Observation:**\
The heatmap shows, within each education group, the proportion of respondents in each income quartile (Q1 = lowest income, Q4 = highest). Respondents with primary and lower secondary education are more concentrated in the lower income bands (Q1–Q2), while upper secondary and “others” show a relatively higher share in the upper bands (Q3–Q4). This pattern suggests a mild positive relationship between education attainment and income.

::::

------------------------------------------------------------------------


# Overall Observations 
Income of informal rural labourers in the northern mountainous regions of Vietnam is highly uneven and right-skewed, with most labourers clustered at lower income levels and a small group earning much more.

Across the factors examined, education stands out as the strongest and most consistent correlate of income, reflected in higher medians and a greater share in upper income quartiles for more educated respondents.

Credit access and vocational training type show secondary, weaker patterns , but dispersion remains large, suggesting other influences are also important. In contrast, province, gender, savings , and PPO1 show modest or minimal differentiation in income.

